---
bibliography: [rmd/references.bib, bib/thesis.bib]
link-citations: true
---

<!-- \FloatBarrier -->

```{r setup, include=FALSE}
# knitr::opts_chunk$set(dpi=300)
```

# Bioinformatics methods {#methods .unnumbered}

## Version controlled data analysis using git {#m1 .unnumbered}

Version control (also known as source control or revision control) is a
software engineering process that involves managing changes to a
computer program or document using a version control system (VCS). In
graph theory, revisions form a directed acyclic graph that represents
the growth line. The revisions to a computer program occur over time and
can be tracked using revision numbers, timestamps, or descriptive text.
While file copies have been used in the past to provide some form of
version control, they are error-prone and time-consuming. The file is
typically copied to another location, and unless the files are cleverly
time-stamped, this simple approach may be error-prone.

(ref:mfc1) \textbf{Overview of version control system (VCS).} Adapted
from [@chacon2021].

```{r mf1, echo=FALSE, fig.align='center', fig.cap="(ref:mfc1)", fig.scap="Overview of version control system", fig.width = 4.96, fig.height=5.6}
img <- readPNG("figure/methods/in/1.png")
grid.raster(img)
```

Programmers developed a local VCS with a database that records all
modifications to the files under revision control to keep track of the
versions. A local VCS is an example of a revision control system (RCS).
Since it is difficult to collaborate with a local VCS, a centralized VCS
was developed, such as Apache Subversion (SVN). All changes to files are
monitored under the centralized server using Centralized VCS. A
centralized VCS, on the other hand, has significant drawbacks. No one
can connect or save versioned changes to something they are working on
if the server goes down for an hour. This leads to the development of
the distributed VCS, such as git. In a distributed VCS, the repository
can be completely mirrored, including the entire version history. A
schematic of distributed VCS is shown in Figure \ref{fig:mf1}, adapted
from [@chacon2021].

GitHub ([github.com](https://github.com)) a distributed version control
system, was used to manage data analysis projects and software
development projects throughout this doctoral thesis work. One downside
of using GitHub is that files greater than 50 megabytes cannot be
hosted, which means that large files such as sequencing data, FASTQ
files, and BAM files are not version managed. To solve this issue,
git-fat was used, a version-controlled tool that offloads files to a
local directory or another server [@brown2018].

## Data analysis directory organization {#m2 .unnumbered}

ScienceCloud service of the University of Zurich was used for computing.
On ScienceCloud, a data volume is mounted on my instance (virtual
computer). In the data volume, two directories were created, namely
"Data" and "Project". The "Data" repository stores raw sequencing files,
FASTQ files, and "gitfat" objects, while the Project directory stores
all analysis and code and is version controlled with git. A schematic
representation of this workflow can be seen in Figure \ref{fig:mf2}.
This is done for each dataset analysed.

(ref:mfc2) \textbf{Organization of data analysis projects.}

```{r mf2, echo=FALSE, fig.align='center', fig.cap="(ref:mfc2)", fig.scap="Organization of data analysis projects", fig.width = 5.25, fig.height=2.98}
img <- readPNG("figure/methods/in/2.png")
grid.raster(img)
```

This organization of the data analysis projects ensures open and
reproducible research. As all the analyses are version controlled, one
can look at the analysis conducted back in time. Also, each analysis
script has a log file, which stores the information of the tools version
and any warnings or errors encountered during the analyses. Further, at
the end of the project, the GitHub repository can be made public and all
the code for intermediate data analysis are available to the research
community.

## Pipelines for data analysis {#m3 .unnumbered}

I worked on multi-omics datasets for my doctoral research, which led to
the development of data analysis pipelines for RNA sequencing (RNA-seq),
Assay of Transposase Accessible Chromatin sequencing (ATAC-seq),
Whole-genome bisulfite sequencing (WGBS), Reduced representation
bisulfite sequencing (RRBS), Chromatin immunoprecipitation followed by
sequencing (ChIP-seq), and short RNA sequencing (sRNA-seq).

### Quality check and trimming of sequencing datasets {#m3.1 .unnumbered}

The goal of data quality check and trimming is to find and clean any
data quality issues in the dataset and possibly account for them. The
sequencer determines the nucleotide bases in a DNA or RNA library during
sequencing. A small sequence, known as a read, is created for each
fragment in the library, which is a sequence of nucleotides. In a single
experiment, sequencing technologies, such as Illumina, can generate a
large number of sequence reads. The first step after receiving the
sequenced data is to assess the data quality to determine the
reliability of the sequenced reads. The FASTQ files are examined using
`FastQC` [@andrews2012]. `FastQC` provides a comprehensive view of the
data, including the number of sequenced reads, base sequence quality,
"N" content (if a sequencer cannot call a base, due to the technical
limitations, it enters "N"; of each sequencing platform.), sequence
duplication levels, and adapter information. A Hypertext Markup Language
(HTML) report generated by the software, for each sample, helps to
understand the pre-processing steps required. The quality assessment is
done using the quality scores (Q-scores) in the FASTQ files, which are
translated to quality statistics in `FastQC` reports and plotted as
boxplots. Q score is defined as the base-calling error probabilities and
is calculated by the formula: `**Q = -log10 (P)**`. For instance, if a
nucleotide base is assigned a Q score of 30, this is equivalent to the
probability of an incorrect base call of 1/1000 times, which means that
the accuracy of base calling is 99.9%. A Q score of 30 represents
perfect base calling with no errors and ambiguities and is considered a
benchmark for quality in high-throughput sequencing (HTS) [@ewing1998].

(ref:mfc3)
\textbf{Workflow for quality check and quality control for sequencing data.}
QC: Quality Control.

```{r mf3, echo=FALSE, fig.align='center', fig.cap="(ref:mfc3)", fig.scap="Workflow for quality check and quality control for sequencing data", fig.width = 4, fig.height=3.9}
img <- readPNG("figure/methods/in/3.png")
grid.raster(img)
```

Adaptors are artificial DNA oligonucleotides and are required for
sequencing by platforms like Illumina. During the process of Illumina
library preparation, adaptors are ligated to the short DNA sequences.
Because of the adapters (attached at the end of reads), the sequencing Q
score is usually low, towards the end of the reads. Therefore, these are
required to be removed before downstream processing of the data.
Furthermore, the sequenced DNA reads could have trailing and leading 'N'
(if a base caller fails to call a base at a genomic location, it will
put 'N' instead of 'A', 'T', 'G' or 'C'). Also, there is a chance that
although long reads were sequenced, small reads appeared after
sequencing, or they became shorter after removal of trailing or leading
'N'. It is always better to get rid of small reads below a certain
length, as they would be mapped to multiple locations and will lead to
misinterpretation. Trimming is the process of modifying the ends of
reads. Trimming can help to increase the number of reads that the
aligner or assembler can successfully use, lowering the amount of
unmapped or unassembled reads. `TrimGalore` is used to trim adapters and
inferior ends of reads with a Phred score of less than 30 (`-q 30`), to
remove Ns from both sides of reads (`--trim-n`), and to discard trimmed
reads that are too short to be informative (e.g. `--length 30`, or `15`
for small RNA-seq data). After this, all results are used to generate an
overview report with `multiQC` [@ewels2016], which is a reporting tool
that parses summary statistics from results of other bioinformatics data
analysis tools. Finally, it is critical to recognize, identify, and rule
out issues that may affect downstream analysis interpretation.

The trimmed data is then aligned to the reference genome, indexed with
the specific tool needed for alignment. For example, for aligning WGBS
data with the reference genome, the reference genome is first indexed
with `Bismark`. This workflow is shown in Figure \ref{fig:mf3}.

Some library preparation kits use the unique molecular identifier (UMI),
which are complex indices and are added to the sequencing libraries
before the polymerase chain reaction (PCR) amplification step. UMIs were
first implemented in the iCLIP protocol [@könig2010] but has been
implemented for RNA-seq, small RNA-seq, single-cell sequencing,
ChIP-seq, and whole-genome sequencing (WGS).

### RNA sequencing {#m3.2 .unnumbered}

RNA sequencing (RNA-seq) is a technique that uses high-throughput
sequencing (HTS) to investigate the quantity and sequences of RNA. It
examines the transcriptome to determine, which of our DNA-encoded genes
are turned on or off, and to what degree. RNA-seq pipeline schematic is
shown in Figure \ref{fig:mf4}.

(ref:mfc4) \textbf{Workflow for RNA sequencing data analysis.}

```{r mf4, echo=FALSE, fig.align='center', fig.cap="(ref:mfc4)", fig.scap="Workflow for RNA sequencing data analysis", fig.width = 6.5, fig.height=3.5}
img <- readPNG("figure/methods/in/4.png")
grid.raster(img)
```

#### Pseudo-alignment and alignment {#m3.2.1 .unnumbered}

Sequence alignment is used to determine where the sequences are similar
to the reference genome and how similar they are. We can estimate where
a read came from by aligning or "mapping" it to a reference genome or
transcriptome. There are two families of methods for mapping reads in
RNA-seq data: alignment and pseudo-alignment. For alignment, aligners
such as `STAR` or `Rsubread` are used. They align reads to a genome or
transcriptome and perform spliced alignment. They return the reads'
base-level alignments. Transcript quantification tools such as `Salmon`
[@patro2017] and `Kallisto` [@bray2016] instead perform
pseudo-alignment. These tools use statistical inference to determine
transcript abundances by mapping reads to the transcriptome. Alignments
from the aligners can then be used for a variety of purposes, such as
for feeding transcript assembly tools, for variant calling pipelines, or
for transcript quantification tools.

Quality-controlled reads were pseudo-aligned using `Salmon` with
automatic detection of the library type (`-l A`), correcting for
sequence-specific bias (`--seqBias`), and correcting for fragment GC
bias correction (`--gcBias`) on a `transcript index` prepared from
GENCODE, with additional piRNA precursors and transposable elements
(concatenated by family) from Repeat Masker as in [@gapp2020].
Downstream analyses including differential analysis; differential
transcript usage and differential isoform switching; and functional
analyses, are performed on the pseudo-alignment from `Salmon`.

Additionally, alignment is performed using the `Rsubread` package
[@liao2019] using the `subjunc()` function, specific for RNA-seq data.
Analysis of duplicates, differential exon usage, and differential 3'
usage are performed on alignment obtained from `Rsubread`.

A direct comparison of genome alignment with transcriptome
pseudo-alignment was performed in [@yi2018] and the authors found that
both approaches produce similar quantifications. However, [@wu2018]
suggested that it is not optimal to use alignment-free methods to
analyze and quantify lowly expressed genes and small RNAs.

#### Duplicated reads analysis {#m3.2.2 .unnumbered}

Duplicated reads are different copies of the exact same sequence. Unless
a genomic region is highly expressed, most mRNA fragments are expected
to be unique. A low degree of repetition may indicate that the target
sequence is well-covered, whereas a high level of duplication is more
likely to signal bias. Duplicate reads might be caused by PCR
duplication or legitimately overrepresented sequences. Because PCR
amplification is more efficient for some sequences than others, it can
misrepresent the true proportion of sequences in the input, whereas
truly overrepresented sequences are the result of very abundant
transcripts in an RNA-Seq library and are an expected case and not of
concern because they accurately represent the input.

Throughout the research, duplicated reads were marked in the aligned
data using `sambamba` [@tarasov2015], and further, the quality issues
with PCR duplicates are analyzed using `dupRadar` [@sayols2016].
`dupRadar` aids in separating the fraction of readings produced by
artefacts from the fraction resulting from high expression.

#### Differential expression analysis {#m3.2.3 .unnumbered}

Differential expression analysis (at the transcript or gene level) is
the statistical examination of normalised read count data to find
quantitative differences in expression levels between experimental
groups.

Read counts at the transcript level are obtained from pseudo-aligned
data and are aggregated at the gene level. Low counts of candidate
genes/transcripts across samples give minimal evidence for differential
expression. When evaluating false discovery rates, they also add to the
multiple testing burden, lowering the ability to find differentially
expressed genes. A pre-filtering of counts is therefore performed using
the `filterByExpr()` function from `edgeR` [@robinson2010] with a design
matrix (matrix containing data about multiple characteristics of several
samples, such as biological group, age, and batch for library
preparation) and requiring at least 20 counts (min.counts = 20)
[@germain2020]. Next, the counts are normalized for different sequencing
depths between samples and to eliminate composition biases between
samples, for example, if there are a few highly expressed genes
dominating in some samples, leading to fewer reads from other genes.
Normalization factors are obtained using the `TMM` normalization method
[@robinson2010] from the `edgeR` package, using the calcNormFactors
function. `TMM` stands for Trimmed Mean of M values, in which the counts
for the samples are scaled using a weighted trimmed mean of the log
expression ratios. Differential expression analysis at gene and
transcript level is performed using limma-voom [@law2014] pipeline from
limma [@ritchie2015] package.

#### Exploratory data analysis {#m3.2.4 .unnumbered}

Exploratory data analysis of counts (using principal component analysis,
PCA), differential analysis results (volcano plot, MA plot, and
heatmap), and assessment of library composition is performed using
`plgINS`, a versatile `R` package,
([github.com/ETHZ-INS/plgINS](https://github.com/ETHZ-INS/plgINS)),
under development by Dr. Pierre-Luc Germain. The package also comes with
a shiny application.

#### Differential transcript usage and differential isoform switching {#m3.2.5 .unnumbered}

Differential transcript usage (DTU) analysis looks for proportional
differences in the makeup of a gene's transcripts between conditions.

DTU is performed using `IsoformSwitchAnalyzeR` [@vitting-seerup2017;
@vitting-seerup2019]. Isoforms are annotated via integration of a wide
range of (predicted) annotations: Pfam was used for prediction of
protein domains, CPC2 (version 2.0) for calculation of the coding
potential, SignalP (version 5.0) for prediction of Signal Peptides, and
IUPred2A was used to predict Intrinsically Disordered Region (IDRs) and
Intrinsically Disordered Binding Regions (IDBRs). The results from Pfam,
CPC2, IUPred2A, and SignalP are used with `IsoformSwitchAnalyzeR` for
annotation and prediction of functional consequences for the identified
isoform changes.

#### Differential exon usage and differential 3' untranslated regions usage {#m3.2.6 .unnumbered}

Differential exon usage (DEU) analysis looks for variations in exon
usage between experimental conditions. mRNAs' 3' untranslated regions
(3' UTRs) are best recognized for regulating mRNA-based functions such
as mRNA localization, stability, and translation [@mayr2019].
Differential UTR usage looks for variations in 3' UTR usage between
experimental conditions.

DEU and differential 3' UTR usage are performed using the diffUTR R
package [@gerber2021].

#### Functional analysis {#m3.2.7 .unnumbered}

The output of RNA-seq differential expression analysis is a list of
significant differentially expressed genes (DEGs). The goal of
functional analysis is to provide biological insight of DEGs. Functional
analysis on Kyoto Encyclopedia of Genes and Genomes (KEGG)
[@kanehisa2021], Reactome [@jassal2020], Gene Ontology (GO)
[@consortium2021] and curated pathway list from the Gene Set Enrichment
Analysis (GSEA) [@subramanian2005] databases is performed using
multiGSEA
([github.com/lianos/multiGSEA](https://github.com/lianos/multiGSEA)),
which facilitates the analysis with GOseq [@young2010], fGSEA
[@korotkevich2016], Correlation Adjusted MEan RAnk gene set test
(CAMERA) and camera pre-ranked (CAMERA-PR) [@wu2012], ROtation testing
using MEan Ranks (ROMER) and fry [@ritchie2015], and ROtAtion gene Set
Testing (ROAST) [@wu2010]. An advantage of using multiGSEA is that it
has standardized input and output and provides a Shiny app for
exploratory data analysis
([github.com/lianos/multiGSEA.shiny](https://github.com/lianos/multiGSEA.shiny)).
With multiGSEA, it is easy to specify to use testing relative to a
threshold (TREAT) method [@mccarthy2009], which takes a user-specified
log fold change cut-off and recalculates the moderated t-statistics and
p-values. Using TREAT only affects enrichment tests that first threshold
the genes in the experiment as "significant" or not, like GOseq and not
tests like the camera.

### Assay of Transposase Accessible Chromatin sequencing {#m3.3 .unnumbered}

The Assay of Transposase Accessible Chromatin sequencing (ATAC-seq) is
widely used in studying chromatin biology to study chromatin
accessibility and to identify open chromatin regions (OCRs). The
nucleosome is the most basic component of chromatin. Each nucleosome is
made up of about two turns of DNA wrapped around a set of eight proteins
called histones known as histone octamer. With the help of nucleosomes,
the genome is securely packed and structured, known as chromatin. The
organization and accessibility of DNA are influenced by a number of
factors, including chromatin structure, nucleosome location,
transcription factors, and histone modifications. As a result, these
factors have a role in the activation and inactivation of genes. The
genome is treated with a hyperactive version of the Tn5 transposase to
insert sequencing adapters into OCRs [@buenrostro2015]. Because it is
easier, faster, and uses fewer cells than competing approaches like
FAIRE-Seq and DNase-Seq, ATAC-Seq has become popular for discovering
accessible regions of the genome. The schematic of the ATAC-seq pipeline
is shown in Figure \ref{fig:mf5}.

(ref:mfc5) \textbf{Workflow for ATAC sequencing data analysis.} QC:
Quality Control, NFF: Nucleosome Free Fragments.

```{r mf5, echo=FALSE, fig.align='center', fig.cap="(ref:mfc5)", fig.scap = "Workflow for ATAC sequencing data analysis", fig.width = 3.55, fig.height=7}
img <- readPNG("figure/methods/in/5.png")
grid.raster(img)
```

#### Alignment, post-alignment quality control, and peak calling {#m3.3.1 .unnumbered}

Alignment of the ATAC-seq data is performed using `Bowtie2`
[@langmead2012] with the following parameters: fragments up to 2 kb were
allowed to align (-X 2000), entire read alignment (--end-to-end),
suppressing unpaired alignments for paired reads (--no-mixed),
suppressing discordant alignments for paired reads (--no-discordant) and
minimum acceptable alignment score with respect to the read length
(--score-min L,-0.4,-0.4). Tn5 inserts adapters separated by 9bp when it
cuts an accessible chromatin region [@kia2017]. This indicates that
reads on the positive strand should be adjusted 4 bp to the right and
reads on the negative strands should be shifted 5 bp to the left in
order for the read start site to reflect the centre of where Tn5 binds
as done in [@buenrostro2013], which is important for the footprint
analysis. Using alignmentSieve from deepTools [@ramirez2016], aligned
data (BAM files) are adjusted for the read start sites to represent the
centre of the transposon cutting event (--ATACshift), and filtered for
reads with a high mapping quality (--minMappingQuality 30). In addition,
because the mitochondrial genome is devoid of chromatin packaging
[@bogenhagen2012] a large number of mitochondrial reads can be a concern
in ATAC-seq. Some ATAC-Seq samples have been found to include 80%
mitochondrial reads, prompting the development of wet-lab approaches to
address the problem [@corces2017]. Also, because the open chromatin
regions of interest are typically found in the nuclear genome,
mitochondrial reads are typically excluded from the analysis. Hence,
reads mapping to the mitochondrial chromosome and ENCODE blacklisted
regions (regions where genome assembly results in erroneous signal) were
filtered using alignmentSieve. Because ATAC-seq does not require
rigorous size selection during library preparation, it can identify
nucleosome positions using fragments representing nucleosome-free
fragments (NFFs) nucleosome monomers, and multimers. There is
approximately 147 bp of DNA wrapped around a nucleosome, and in order to
obtain NFFs, which are indicative of transcription factor bindings, the
fragments less than 147 bp must be extracted. To extract NFFs, all
aligned files are merged within experimental groups, reads are sorted by
left-most coordinates and indexed using SAMtools [@li2009], and NFFs
were obtained by selecting alignments with a template length between 40
and 140 inclusively [@jung2017; @goodnight2019]. Peak calling is a
statistical approach that leverages data coverage properties to identify
locations that are enriched as a result of protein/ transcription factor
binding. To find regions corresponding to potential OCRs in the
nucleosome-free regions, we want to identify regions where reads have
piled up greater than the background read coverage. Peak calling on the
NFFs is performed using MACS2 [@zhang2008] with genome size in base-pair
(-g genomeSize) and specifying the paired-end BAM file format (-f
BAMPE).

#### Peaks annotation and differential accessibility analysis {#m3.3.2 .unnumbered}

The goal of peak annotation is to map peaks to the regulatory elements.
The peaks are annotated with transcript, and the distance to the nearest
transcription start site, based on overlap with gene transfer format
(GTF) file obtained from the GENCODE. The number of extended reads
overlapping in the peak regions is calculated using the csaw package
[@lun2015]. Peak regions that did not have at least 15 reads in at least
40% of the samples were filtered out. Normalization factors were
obtained on the filtered peak regions using the `TMM` normalization
method [@robinson2010] and differential accessibility analysis is
performed using the Genewise Negative Binomial Generalized Linear Models
with Quasi-likelihood (glmQLFit) test from the `edgeR` package. `edgeR`
has recently been benchmarked and recommended for differential chromatin
accessibility analysis [@gontarz2020].

#### Functional analysis {#m3.3.3 .unnumbered}

GO analysis is performed on differentially accessible regions (DARs)
with the rGREAT package
([github.com/jokergoo/rGREAT](https://github.com/jokergoo/rGREAT)),
which is a wrapper around the Genomic Regions Enrichment of Annotations
Tool (GREAT) [@mclean2010]. Transcription factor motif enrichment
analysis is performed using the marge package [@amezquita2018], which is
a wrapper around the Homer tool [@heinz2010].

#### Differential accessibility analysis at transposable elements {#m3.3.4 .unnumbered}

Transposable elements (TEs) GTF file is obtained from
[http://labshare.cshl.edu/shares/mhammelllab/www-data/TEtranscripts/TE_GTF/mm10_rmsk_TE.gtf.gz](http://labshare.cshl.edu/shares/mhammelllab/www-data/TEtranscripts/TE_GTF/mm10_rmsk_TE.gtf.gz%20on%2003.02.2020).
The GTF file provides hierarchical information about TEs: **Class**
(level 1, eg. LTR), **Family** (level 2, eg. LTR L1), **Subtype** (level
3, eg. LTR L1 L1_Rod), and **Locus** (level 4, eg. LTR L1 L1_Rod
L1_Rod_dup1). TE loci are annotated based on overlap with GENCODE as
described above for ATAC-seq peaks. Filtered BAM files (without reads
mapping to blacklisted or mitochondrial regions) are used for analyzing
TEs. Mapped reads are assigned to TEs using featureCounts from the R
package `Rsubread` [@liao2019] and were summarized to Subtypes (level
3), allowing for multi-overlap with fractional counts, while ignoring
duplicates. The number of extended reads overlapping at the TE loci is
obtained using the csaw package [@lun2015]. Subtypes that did not have
at least 15 reads, and loci that did not have at least 5 reads in at
least 40% of the samples, were filtered out. Normalization and
differential accessibility analysis is performed as described above for
the peaks. GO and motif enrichment analysis is performed as described
above for the peak regions.

### Chromatin immunoprecipitation followed by sequencing {#m3.4 .unnumbered}

A substantial part of gene expression regulation is controlled by
protein-DNA interactions. Proteins like transcription factors and
histones influence how much and where genes are expressed. ChIP,
followed by sequencing, can be used to investigate the interaction
between a DNA sequence and a protein. There are two types of ChIP:
cross-linked ChIP (X-ChIP) and native ChIP (N-ChIP). Formaldehyde is
utilized to crosslink histones to DNA, and ultrasound sonication is used
to fragment the DNA in X-ChIP, whereas the native covalent interaction
between protein and DNA is exploited in N-ChIP [@park2009]. The
fundamental purpose of ChIP is to map global binding sites for any
protein of interest, such as transcription factors (TF) and histones, in
order to investigate specific modifications. In vivo, the approach
generates a library of DNA locations bound to the protein of interest.
Following ChIP on samples, library preparation, PCR amplification, and
ultimately deep sequencing are performed. The schematic of the ChIP-seq
pipeline is shown in Figure \ref{fig:mf6}.

(ref:mfc6) \textbf{Workflow for ChIP sequencing data analysis.} QC:
Quality Control.

```{r mf6, echo=FALSE, fig.align='center', fig.cap="(ref:mfc6)", fig.scap="Workflow for ChIP sequencing data analysis", fig.width = 4, fig.height=6}
img <- readPNG("figure/methods/in/6.png")
grid.raster(img)
```

#### Alignment and peak calling {#m3.4.1 .unnumbered}

Alignment is performed using `Bowtie2`. Reads with more than 3
mismatches are removed from the aligned data, as suggested in
[@royo2016], and reads with low mapping quality (--minMappingQuality 30)
or mapping to the mitochondrial chromosome or aforementioned blacklisted
regions are also filtered out. Peak calling is performed using MACS2
with mouse genome size in base-pair (-g genomeSize) and the BAM file
format (-f BAM/PEBAM), together with the ChIP input sample.

#### Peaks annotation, differential analysis, and functional analysis {#m3.4.2 .unnumbered}

These analyses are performed as described above for the ATAC-seq peaks.

### Bisulfite sequencing {#m3.5 .unnumbered}

DNA methylation (DNAme) involves the addition of methyl groups to the
DNA molecule. It most normally occurs at the C5 position of cytosines
inside CpG dinucleotides, where it is stable during mitosis and meiosis.
Methylation can change the activity of DNA without changing its
sequence. DNAme is a reversible, cell-type-specific DNA modification
that is usually persistent through cell division. It is not equally
distributed across the genome, but rather is linked to CpG density.
High-throughput bisulfite sequencing is one of the most reliable methods
for measuring DNA methylation. In bisulfite conversion, unmethylated
cytosine residues are converted to uracil, but methylated cytosines
remain intact [@frommer1992], allowing methylation and unmethylated
cytosine bases to be distinguished. Following PCR amplification, these
uracils are converted to thymines. This is followed by the WGBS. WGBS is
considered a gold-standard technology for DNA methylation detection as
it provides a qualitative, quantitative, and efficient technique to
identify 5-methylcytosine at single base-pair resolution. The cost of
sequencing an entire genome is typically quite high. As a result,
despite its use on huge genomes like the human genome, significant
numbers of individual samples are rarely sequenced. For this reason,
RRBS was developed, in which the bisulfite process occurs but only
around 1% of the genome is sequenced. This makes it possible to sequence
the genomes in greater numbers. The idea behind RRBS is to enrich the
sequencing library with CpG-dense sections of the genome to boost the
sequencing depth of cytosine-rich areas [@meissner2005]. This is done by
using CpG-specific restriction enzymes like MspI, which identifies the
sequence 5'-CCGG-3'. MspI cleaves the complete genome in RRBS, leaving
at least two cytosines per cleaved DNA fragment (one cytosine at either
end of the MspI-MspI fragment). Sequencing libraries are made from
size-selected MspI DNA fragments, followed by bisulfite conversion and
sequencing of the converted library. The schematic of the BS-seq
pipeline is shown in Figure \ref{fig:mf6}.

(ref:mfc7) \textbf{Workflow for Bisulfite sequencing data analysis.} QC:
Quality Control, DML: differentially methylated loci, DMR:
differentially methylated regions, CNV: copy-number variation.

```{r mf7, echo=FALSE, fig.align='center', fig.cap = "(ref:mfc7)", fig.scap="Workflow for Bisulfite sequencing data analysis", fig.width = 6.47, fig.height=2.83}
img <- readPNG("figure/methods/in/7.png")
grid.raster(img)
```

#### Alignment and methylation calling {#m3.5.1 .unnumbered}

Mismatches between the reads and the reference genome are introduced
during bisulfite conversion, resulting in slow and imprecise mapping.
Further, the DNA code's complexity is reduced. Hence, specialized tools
are required for aligning the bisulfite sequencing data. Alignment of
the bisulfite sequencing data is performed using `Bismark`
[@krueger2011] on a genome index built using
`Bismark`\_genome_preparation. Methylation information for individual
cytosines (methylation calls) is extracted using the
`Bismark`\_methylation_extractor tool from the `Bismark` package.

#### Differential methylation analysis {#m3.5.2 .unnumbered}

Differential analysis can be performed at the loci level (differentially
methylated loci, DML) or at the region's level (differentially
methylated regions, DMRs). For the identification of DML and DMRs, the
data is first filtered by the coverage among samples. DML are identified
using `edgeR`. Identification of DMRs is not so straightforward, as
defining regions boundaries is not easy the tool has to take into
account correlation across loci of a region. Hence, during a course, we
benchmarked different tools for detecting DMRs
[dktanwar.github.io/sta426-project-dmr-comparison](https://dktanwar.github.io/sta426-project-dmr-comparison/report/project_report.html)
and found dmrseq to outperform all other methods that we compared. DMRs
from RRBS data are identified using the parameters recommended by Dr.
Keegan Korthauer**,** author of dmrseq
([github.com/kdkorthauer/dmrseq/issues/14](https://github.com/kdkorthauer/dmrseq/issues/14)).

#### Copy number variation analysis {#m3.5.3 .unnumbered}

The phenomenon of copy number variation (CNV) occurs when regions of the
genome are repeated and the amount of repeats differs between
individuals. CNV is a type of structural variation that occurs when a
duplication or deletion event occurs that impacts a large number of base
pairs. To infer and visualize CNV from WGBS data, we use the CNVkit
[@talevich2016].

#### DML and DMR annotation, and functional analysis {#m3.5.4 .unnumbered}

These analyses are performed as described above for the ATAC-seq data.

### Short RNA-seq {#m3.6 .unnumbered}

The short RNA-seq pipeline is not described in this section, as a method
for analyzing small RNA-seq data has been developed and is discussed in
detail in Chapter 3.

### Multi-omics analysis {#m3.7 .unnumbered}

Using different experiments, followed by sequencing, on the same samples
or cells/tissues, such as RNA-seq, ATAC-seq, ChIP-seq, and BS-seq,
results in multi-dimensional omics datasets that can be used to
investigate relationships between different biological processes, such
as gene expression, chromatin accessibility, histone modifications, and
DNA methylation. Integrating multi-omics datasets is most challenging
and is not straightforward to perform, as there are no standard recipes.

We present a straightforward and biologically useful approach for
performing multi-omics analysis on RNA-seq, ATAC-seq, ChIP-seq, and
BS-seq datasets. First, we overlap the DARs with ChIP-seq peaks
enrichment, RNA-seq expression of overlapping genes with DARs, and
methylation values of DARs overlapping loci from BS-seq data into one
matrix, overlapMatrix. The DARs are initially divided into proximal
(less than +/- 2.5 kb from a TSS) and distal (more than +/- 2.5 kb from
a TSS) regions, as per the ENCODE guidelines [@thurman2012;
@harrow2012]; the proximal regions are further divided into active and
inactive (proximal active and proximal inactive) groups based on nearby
gene expression (if nearby gene expression is detected) (**step 1**).
Next, these three groups are separated based on whether the biological
groups are more or less accessible (**step 2**). This classification
would reveal six separate clusters of DARs, each with its own biological
profile. Further, proximally active genes are categorised based on their
up/ down expression (**step 3**). Moreover, we categorised DARs
depending on whether they had H3K4me3, which is a critical marker for
the active transcription and developmentally stable genes [@fu2019];
H3K27ac, which is associated with the higher activation of transcription
and therefore defined as an active enhancer mark [@zhang2020]; or dual
H3K4me3/H3K27ac enrichment [@beacon2021]; as well as H3K27me3
enrichment, which is involved in transcriptional silencing
[@bogliotti2012] (**step 4**). The schematic is shown in Figure
\ref{fig:mf8}. Furthermore, gene ontology and TF motif enrichment
analysis can be done on these categorised DARs. For a more in-depth
understanding, all of the regions can be further classified based on
their methylation state (high or low). Please see [Chapter 2](#chapter2)
for an application of this multi-omics analysis.

```{=tex}
\newpage
\begin{landscape}

(ref:mfc8) \textbf{Workflow for multi-omics data integration.}

```{r mf8, echo=FALSE, fig.align='center', fig.cap="(ref:mfc8)", fig.scap="Workflow for multi-omics data integration", fig.width = 9, fig.height=7}
img <- readPNG("figure/methods/in/8.png")
grid.raster(img)
```
\end{landscape}
```
Visualization of these regions is an essential step. We used
EnrichedHeatmap [@gu2018] and ComplexHeatmap [@gu2016] packages for the
visualization of categorized DARs.

**ATAC-seq:** ATAC-seq data is the starting point and we first find the
middle of each DARs. After that, we obtain the coverage for 1kb upstream
and downstream from the middle of DARs. We use the `normalizeToMatrix`
function from the EnrichedHeatmap package and normalized the signal
using the weighted mean between the intersected parts and un-intersected
parts(mean_mode = "w0") in a window of 50bp. In the normalized matrix,
each row corresponds to DAR and each column corresponds to a window
either on upstream or downstream of middle of the peak.

$$
v_{w 0}=\frac{\sum_{i=1}^{n} x_{i} w_{i}}{W+W^{\prime}}
$$

${W}$ is the sum of width of the intersected parts
$\left(\sum_{i}^{n} w_{i}\right)$ and $W^{\prime}$ is the sum of width
for the non-intersected parts.

For visualization, DARs are ordered by the enriched scores. For each DAR
in the matrix, values are denoted as ${x}$, indices as
$1, \ldots, n_{1}$ for upstream windows, indices as $n_{1}+1, \ldots, n$
for downstream windows, and $n_{2}=n-n_{1}$ are the enriched score that
are calculated as the sum of ${x}$ weighted by distance to middle of the
peak. DAR has a higher enriched score when there is more signal focused
on it.

$$
\sum_{i=1}^{n_{1}} x_{i} \cdot i / n_{1}+\sum_{i=n_{1}+1}^{n} x_{i} \cdot(n-i+1) / n_{2}
$$

Moreover, we also plot the log\textsubscript[2]fold-changes of DARs using
ComplexHeatmap.

**ChIP-seq:** ChIP-seq data coverage for 1kb upstream and downstream
from the mid of DARs is obtained. We use the `normalizeToMatrix` function
from the EnrichedHeatmap package and normalized the signal using the
weighted mean between the intersected parts and un-intersected parts
(mean_mode = "w0") in a window of 50bp.

**BS-seq:** BS-seq data methylation values for 1kb upstream and
downstream from the mid of DARs is obtained. We use the
`normalizeToMatrix` function from the EnrichedHeatmap package and
normalized the signal using the mean of all the signals (mean_mode =
"absolute") in a window of 50bp.

$$
v_{a}=\frac{\sum_{i=1}^{n} x_{i}}{n}
$$

**RNA-seq:** For RNA-seq data, we used ComplexHeatmap to plot the
log\textsubscript[2]-fold-changes of expression values between the groups, and also the
log\textsubscript[2]-fold-changes obtained after the differential analysis.
